---
title: "Predicting Treatment Slopes with Pre-Treatment Clinical and Gene Expression Data"
output:
  html_document:
    df_print: paged
---

# Load libraries and custom code

Here we will load all the R libraries necessary and all the custom code as well (e.g., functions for running the LASSO models, making spaghetti plots, etc)

```{r, warning=FALSE, message=FALSE}
# load libraries and other custom code
easypackages::libraries("readxl","ggplot2","patchwork",
                        "here","limma","reshape2","glmnet",
                        "genefilter","Biobase","psych","nlme")
source(here("code","lasso_functions.R"))
source(here("code","spaghettiPlot.R"))
source("/Users/mlombardo/Dropbox/R/get_ggColorHue.R")

# number of permutations to set for all the permutation tests run down below
nperm = 1000

fontSize = 5
```

# Load treatment data

Load in excel files that contain the longitudinal treatment measure (aSLP).

```{r, warning=FALSE, message=FALSE}
# Read in melted treatment data
txplotpath = here("plots")

txfile = here("data","melted_tx_data.xlsx")
melted_tx_data = read_excel(txfile)

melted_tx_data_orig = melted_tx_data

txfile = here("data","tx_data.xlsx")
tx_data = read_excel(txfile)
tx_data = as.data.frame(tx_data)
rownames(tx_data) = tx_data$subjectId

data2use = data.frame(melted_tx_data)
data2use$subgrp = factor(data2use$subgrp, exclude = "NA")

# run statistical model
y_var = "value"
x_var = "age"
fx_form = as.formula(sprintf("%s ~ 1", y_var, x_var))
rx_form = as.formula(sprintf("~ %s|subjectId",x_var))

ctrl <- lmeControl(opt='optim', msMaxIter = 500)
ml_method = "REML"
mod2use = eval(substitute(lme(fixed = fx_form, random = rx_form, data = data2use,
                                 na.action = na.omit, control=ctrl, method = ml_method),
                             list(fx_form = fx_form, rx_form = rx_form)))
mod2 = mod2use
summary(mod2)
# grab subject-specific intercepts and slopes
sub_spec_int_slopes2 = coef(mod2use)
sub_spec_int_slopes2$subjectId = factor(rownames(sub_spec_int_slopes2))
colnames(sub_spec_int_slopes2)[c(1,2)] = c("rfx_intercept","rfx_slope")

# merge subject-specific intercepts and slopes with treatment data
tx_data = merge(x = tx_data, y = sub_spec_int_slopes2[,c("subjectId","rfx_intercept","rfx_slope")], by.x = "subjectId", by.y = "subjectId")
```

# Load gene expression data

Here we load in the superset of all microarray gene expression data, and then do some cleaning up of things so that we are left with a final subset of subjects in this treatment dataset.

```{r, warning=FALSE, message=FALSE}
# load superset of gene expression data
load("~/Dropbox/data/ucsd_ace_microarray_data/dataQuantile.rda")

# grab pheno data
pheno_data = pData(dataQuantile)
# grab expression data
exprData = exprs(dataQuantile)
# grab meta data for genes
meta_data = fData(dataQuantile)

# subjectId for subset of treatment subjects
subids2use = tx_data$subjectId

# grab pheno data for subset of subjects with expression data
pheno_data_sub = subset(pheno_data,is.element(pheno_data$subjectId,subids2use))
pheno_data_sub$exprColName = rownames(pheno_data_sub)
# find duplicates OR subjects with GEX data older than TX Intake age and remove them
duplicates2remove = c("4807258001_B","4807258003_C","4807258003_A",
                      "4807135001_F","4807135021_B","4807135022_E",
                      "4807135045_E","4807135042_E","4807135042_F","107","243","187","359","272",
                      "319","228","339","133","200")

mask = is.element(pheno_data_sub$exprColName,duplicates2remove)
pheno_data_sub = pheno_data_sub[!mask,]

# tidy up the pheno_data_sub data frame
colnames(pheno_data_sub)[colnames(pheno_data_sub)=="age"] = "gex_age"
pheno_data_sub$sex = factor(pheno_data_sub$sex)
pheno_data_sub$batch = factor(pheno_data_sub$batch)
pheno_data_sub$diagnosis_binary = factor(pheno_data_sub$diagnosis_binary)
pheno_data_sub$subjectId = factor(pheno_data_sub$subjectId)
pheno_data_sub$blooddrawid = factor(pheno_data_sub$blooddrawid)
pheno_data_sub$exprColName = factor(pheno_data_sub$exprColName)

# find RIN for subs in pheno_data_sub
rin_data = read_excel("/Users/mlombardo/Dropbox/data/ucsd_ace_microarray_data/pData_HT12_WG6_processedv2.xlsx",na = "NA")
rin_data_sub = as.data.frame(subset(rin_data,is.element(rin_data$blooddrawid,pheno_data_sub$blooddrawid)))
rownames(rin_data_sub) = rin_data_sub$subjectId
subs_notin_rin_data_sub = pheno_data_sub$subjectId[!is.element(pheno_data_sub$subjectId,
                                                               rin_data_sub$subjectId)]
new_rin_data_rows = data.frame(matrix(nrow=length(subs_notin_rin_data_sub), ncol=ncol(rin_data_sub)))
rownames(new_rin_data_rows) = subs_notin_rin_data_sub
colnames(new_rin_data_rows) = colnames(rin_data_sub)
rin_data_sub = rbind(rin_data_sub, new_rin_data_rows)
rin_data_sub = rin_data_sub[pheno_data_sub$subjectId,]

# insert RIN, subgrp, and rfx_slope into pheno_data_sub
pheno_data_sub$RIN = NA
pheno_data_sub$subgrp = NA
pheno_data_sub$rfx_slope = NA
for (subjectId in pheno_data_sub$subjectId) {
  rin2use = rin_data_sub[subjectId,"Final_RIN"]
  pheno_data_sub[pheno_data_sub$subjectId==subjectId,"RIN"] = rin2use
  
  subgrp2use = as.character(tx_data[tx_data$subjectId==subjectId,"subgrp"])
  pheno_data_sub[pheno_data_sub$subjectId==subjectId,"subgrp"] = subgrp2use
  
  slopes2use = tx_data[tx_data$subjectId==subjectId,"rfx_slope"]
  pheno_data_sub[pheno_data_sub$subjectId==subjectId,"rfx_slope"] = slopes2use
} # for (subjectId in pheno_data_sub$subjectId) {

pheno_data_sub$subgrp = factor(pheno_data_sub$subgrp)

# fill missing RIN with meanRIN
meanRIN = mean(pheno_data_sub$RIN, na.rm=TRUE)
na_rin_mask = is.na(pheno_data_sub$RIN)
pheno_data_sub$RIN[na_rin_mask] = meanRIN 

# re-order expression data columns to be the same order as the rows of pheno_data_sub
exprData_sub = exprData[,as.character(pheno_data_sub$exprColName)]

# write pheno data and expression data to csv files
write.csv(pheno_data_sub, file = here("data","exprPhenoData_Microarray.csv"))
write.csv(exprData_sub, file = here("data","exprData_Microarray.csv"))
write.csv(meta_data, file = here("data","exprMetaData_Microarray.csv"))

# subset tx_data to only those subjects with expression data
tx_data_sub = subset(tx_data, is.element(tx_data$subjectId,pheno_data_sub$subjectId))
rownames(tx_data_sub) = tx_data_sub$subjectId
tx_data_sub$subjectId = factor(tx_data_sub$subjectId)
tx_data_sub$subgrp = factor(tx_data_sub$subgrp)
# rearrange ordering of rows to be same as ordering of 
# rows and columns in pheno_data_sub and exprData_sub
tx_data_sub = tx_data_sub[pheno_data_sub$subjectId,]
# write out file of tx data for only subjects with expression data
write.csv(tx_data_sub, file = here("data","tx_data_exprSubs_Microarray.csv"))

unique_subs2use = unique(tx_data_sub$subjectId)
melted_tx_data = subset(melted_tx_data, is.element(melted_tx_data$subjectId, unique_subs2use))
```

# Preprocess gene expression data

Use variance filtering to remove genes with low expression variance.

```{r, warning=FALSE, message=FALSE}
# read in data
exprPhenoData = read.csv(file = here("data","exprPhenoData_Microarray.csv"))
exprData = read.csv(file = here("data","exprData_Microarray.csv"), row.names = 1)
exprMetaData = read.csv(file = here("data","exprMetaData_Microarray.csv"))

# =============================================================================
# variance filter to grab top X% of genes with largest variance
varianceFilterCutoff = 0.50
filt_exprData = varFilter(as.matrix(exprData), var.cutoff = varianceFilterCutoff)
# grab meta data for filtered gene set
filt_meta_data = meta_data[rownames(filt_exprData),]

# plot the unfiltered dataset
x = melt(as.matrix(exprData))
p = ggplot(data = x, aes(x = value, colour = Var2)) + geom_density() + guides(colour = FALSE)
p

# plot the resulting filtered dataset
x = melt(filt_exprData)
p = ggplot(data = x, aes(x = value, colour = Var2)) + geom_density() + guides(colour = FALSE)
p

# save adjusted expression data
save(exprData,meta_data,exprPhenoData,filt_exprData,filt_meta_data, 
     file = here("data","exprData_Microarray.Rdata"))
```

# Examine if treatment slopes are different as a function of batch, gex age, RIN, or sex 

```{r, warning=FALSE, message=FALSE}
mod2use = lm(formula = rfx_slope ~ batch + gex_age + RIN + sex, data = pheno_data_sub)
anova(mod2use)
```

# Test if age at TX intake correlates with TX slopes 

```{r, warning=FALSE, message=FALSE}
tx_data_sub$TxStartSubgrp = "Before24Mo"
tx_data_sub$TxStartSubgrp[tx_data_sub$age_tx_intake>=24] = "After24Mo"
tx_data_sub$TxStartSubgrp = factor(tx_data_sub$TxStartSubgrp)

# fontSize=20
cols2use = colorRampPalette(c("blue","red"))(100)
p1 = ggplot(data = tx_data_sub, aes(x = age_tx_intake, y = rfx_slope, colour = rfx_slope)) + 
  geom_smooth(method=lm, colour = "black") + geom_point() + scale_color_gradientn(colors=cols2use, name="Treatment Slope") +
  xlab("Age at Treatment Start") + ylab("Treatment Slope") + 
  theme(axis.text.x = element_text(size=fontSize),
        axis.text.y = element_text(size=fontSize),
        axis.title.x = element_text(size=fontSize),
        strip.text.x = element_text(size=fontSize),
        axis.title.y = element_text(size=fontSize))
ggsave(filename = here("plots","age_tx_intake_rfx_slope_correlation_plot_ALL.pdf"))
p1

cor.test(tx_data_sub$age_tx_intake, tx_data_sub$rfx_slope)
```

# Model age at treatment start or treatment start subgroups in LME

```{r, warning=FALSE, message=FALSE}
plotfname2save = file.path(txplotpath,"aSLP_EarlyLateSubgrp.pdf")
data2use = data.frame(melted_tx_data)
tx_data_sub$EarlyLateSubgrp = "Early"
tx_data_sub$EarlyLateSubgrp[tx_data_sub$age_tx_intake>=24] = "Late"
tx_data_sub$EarlyLateSubgrp = factor(tx_data_sub$EarlyLateSubgrp)
early_subids2use = unique(tx_data_sub$subjectId[tx_data_sub$EarlyLateSubgrp=="Early"])
late_subids2use = unique(tx_data_sub$subjectId[tx_data_sub$EarlyLateSubgrp=="Late"])
melted_tx_data$EarlyLateSubgrp = "Early"
melted_tx_data$EarlyLateSubgrp[is.element(melted_tx_data$subjectId,late_subids2use)] = "Late"
melted_tx_data$EarlyLateSubgrp = factor(melted_tx_data$EarlyLateSubgrp)
data2use = merge(melted_tx_data, tx_data_sub[,c("subjectId","days_in_tx")])

# run statistical model
y_var = "value"
x_var = "age"
subgrp_var = "EarlyLateSubgrp"
fx_form = as.formula(sprintf("%s ~ days_in_tx + %s*%s", y_var, x_var, subgrp_var))
rx_form = as.formula(sprintf("~ %s|subjectId",x_var))

ctrl <- lmeControl(opt='optim', msMaxIter = 500)
ml_method = "ML"
mod2use = eval(substitute(lme(fixed = fx_form, random = rx_form, data = data2use, 
                                 na.action = na.omit, control=ctrl, method = ml_method), 
                             list(fx_form = fx_form, rx_form = rx_form)))
anova(mod2use)

# make the plot
lm_res = spaghettiPlot(data2use,
                       x_var = "age",
                       y_var = "value",
                       subgrp_var = "EarlyLateSubgrp",
                       xLabel = "Age (months)",
                       yLabel = "# of Mastered Targeted Skills (aSLP)",
                       modelType = "linear",
                       fname2save = plotfname2save,
                       plot_dots = TRUE,
                       plot_lines = TRUE,
                       ci_band = TRUE,
                       pi_band = FALSE,
                       dot_alpha = 8/10,
                       line_alpha = 8/10,
                       band_alpha = 3/10,
                       xLimits = c(10, 40),
                       yLimits = c(-5,240))

# fontSize = 20
p = lm_res$p
p = p + 
  theme(axis.text.x = element_text(size=fontSize),
        axis.text.y = element_text(size=fontSize),
        axis.title.x = element_text(size=fontSize),
        strip.text.x = element_text(size=fontSize),
        axis.title.y = element_text(size=fontSize))
ggsave(filename = plotfname2save, plot = p)
p_early_late = p
p

# model standard deviation difference between Early vs Late start groups
early_sd = sd(tx_data_sub$rfx_slope[tx_data_sub$EarlyLateSubgrp=="Early"])
late_sd = sd(tx_data_sub$rfx_slope[tx_data_sub$EarlyLateSubgrp=="Late"])
sd_diff = late_sd - early_sd
print(sprintf("Early SD = %f",early_sd))
print(sprintf("Late SD = %f",late_sd))
print(sprintf("SD difference (Late - Early) = %f",sd_diff))

# Run permutation test
num_perm = 10000
vars2use = c("num_perm","perm_early_sd","perm_late_sd","perm_sd_diff")
perm_res = data.frame(matrix(nrow=num_perm, ncol=length(vars2use)))
colnames(perm_res) = vars2use

for (iperm in 1:num_perm){
  set.seed(iperm)
  tx_data_sub$perm_rfx_slope = sample(tx_data_sub$rfx_slope)
  perm_res$perm_early_sd[iperm] = sd(tx_data_sub$perm_rfx_slope[tx_data_sub$EarlyLateSubgrp=="Early"])
  perm_res$perm_late_sd[iperm] = sd(tx_data_sub$perm_rfx_slope[tx_data_sub$EarlyLateSubgrp=="Late"])
  perm_res$perm_sd_diff[iperm] = perm_res$perm_late_sd[iperm] - perm_res$perm_early_sd[iperm]
}

# permutation p-value for whether actual standard deviation difference in real data is bigger than that in permuted data
perm_sd_pval = (sum(perm_res$perm_sd_diff>=sd_diff)+1)/(num_perm+1)
perm_sd_pval 

# make histogram of permuted SD diff and plot actual SD diff as vertical red line
p = ggplot(data = perm_res, aes(x = perm_sd_diff)) + geom_histogram(bins = 100) + geom_vline(xintercept = sd_diff, color="red") + 
  xlab("Standard Deviation Difference (Late - Early)")+ ylab("Count")
p = p + 
  theme(axis.text.x = element_text(size=fontSize),
        axis.text.y = element_text(size=fontSize),
        axis.title.x = element_text(size=fontSize),
        strip.text.x = element_text(size=fontSize),
        axis.title.y = element_text(size=fontSize))
ggsave(filename = here("plots","Early_vs_Late_SD_histogram.pdf"))
p_sd_early_late = p
p
```

# Run LASSO to predict subject-specific slopes with gene expression data

```{r, warning=FALSE, message=FALSE}
load(here("data","exprData_Microarray.Rdata"))
dataType = "Microarray"

USE_FILTERED_DATA = TRUE
if (USE_FILTERED_DATA) {
  data2use = t(filt_exprData)
  metadata2use = filt_meta_data
} else {
  data2use = t(exprData)
  metadata2use = meta_data
} # if (USE_FILTERED_DATA) {
dv2use = exprPhenoData[,"rfx_slope"] 

ZSCORE_EXPRDATA = FALSE
if (ZSCORE_EXPRDATA){
  data2use = scale(data2use)
}

score.LOO.slope = .myLasso.LOO.new(inputData = t(data2use), 
                                   inputTrait = exprPhenoData, 
                                   dataType = dataType)

# show what is the correlation between observed and predicted slopes from the LASSO model.
cor(score.LOO.slope$data2plot)

# print out mean squared error from this model
score.LOO.slope$mse

# Make the spaghetti plot 
tx_data2plot = merge(melted_tx_data, exprPhenoData[,c("subjectId","rfx_slope")], by="subjectId")

lm_res = spaghettiPlot(tx_data2plot, 
                       x_var = "age", 
                       y_var = "value", 
                       subgrp_var = "subgrp", 
                       xLabel = "Age (months)", 
                       yLabel = "# of Mastered Targeted Skills (aSLP)", 
                       modelType = "linear", 
                       fname2save = NULL, 
                       plot_dots = TRUE, 
                       plot_lines = TRUE, 
                       ci_band = FALSE, 
                       pi_band = FALSE, 
                       dot_alpha = 8/10, 
                       line_alpha = 8/10,
                       band_alpha = 3/10, 
                       xLimits = c(10, 40), 
                       yLimits = c(-5,240),
                       lineColor = "rfx_slope",
                       dotColor = "rfx_slope")
single_sub_plot = lm_res$p

# fontSize = 20
long_plot = single_sub_plot
long_plot = long_plot + 
  theme(axis.text.x = element_text(size=fontSize),
        axis.text.y = element_text(size=fontSize),
        axis.title.x = element_text(size=fontSize),
        strip.text.x = element_text(size=fontSize),
        axis.title.y = element_text(size=fontSize))
ggsave(filename = here("plots","aSLPtrajectories_rfx_slope_colored.pdf"), plot = long_plot)


df2plot = data.frame(predSlope = score.LOO.slope$data2plot$predVal, 
                     obsSlope = score.LOO.slope$data2plot$inputTrait, 
                     subjectId = exprPhenoData$subjectId, 
                     subgrp =exprPhenoData$subgrp)

fontSize_tmp = 15
cols2use = colorRampPalette(c("blue","red"))(100)
p2 = ggplot(data = df2plot, aes(x = predSlope, y = obsSlope)) + 
  geom_smooth(method=lm, col="black") + 
  geom_point(aes(colour=obsSlope),size=2) + scale_color_gradientn(colors=cols2use, name = "Slope") +
  xlab("Predicted Slope") + ylab("Actual Slope") + theme(legend.text=element_text(size=fontSize_tmp))+
  theme(axis.text.x = element_text(size=fontSize_tmp),
        axis.text.y = element_text(size=fontSize_tmp),
        axis.title.x = element_text(size=fontSize_tmp),
        strip.text.x = element_text(size=fontSize_tmp),
        axis.title.y = element_text(size=fontSize_tmp))
ggsave(filename = here("plots","gex_lasso_loo_plot2a.pdf"))

p_final = single_sub_plot + p2 + plot_layout(nrow=1,ncol=2)
ggsave(filename = here("plots","gex_lasso_loo_plot2b.pdf"))
p_final
#------------------------------------------------------------------------------

# now lets find the genes that most highly contribute to the model
score.LOO.slope.genesALL = .myLasso.LOO.genes.new(inputData = t(data2use),
                                                  inputTrait = exprPhenoData,
                                                  dataType = dataType)

# find genes of importance
# score.LOO.slope.genesALL = .myLasso.LOO.genes(inputData = data2use, inputTrait = dv2use)
genesALL.keep.LOOslope = as.data.frame(table(unlist(score.LOO.slope.genesALL[1,])))
genesALL.keep.LOOslope = genesALL.keep.LOOslope[with(genesALL.keep.LOOslope, order(-Freq)),]
geneFilter = as.character(genesALL.keep.LOOslope[,1])
good_genes = metadata2use[geneFilter,"geneSymbol"]
res = data.frame(geneSymbol = good_genes, genesALL.keep.LOOslope)
colnames(res)[2] = "ILMN_GENE"
write.csv(res, file = here("results","res_lasso_loo_bestgenes_Microarray2.csv"))
background_list = metadata2use$geneSymbol[is.element(metadata2use$PROBE_ID,colnames(data2use))]
write(background_list, file = here("results","backgroundList_Microarray2.csv"))
res

# permutation test on MSE
null_mse = .myRandFn.LOO.new(inputData = t(data2use),
                             inputTrait = exprPhenoData,
                             nperm = nperm,
                             dataType = dataType,
                             seed2use=123,
                             verbose=FALSE)

# make histogram plot of MSE null distribution
df2plot = data.frame(mse = null_mse)
p = ggplot(data = df2plot, aes(x = mse))
p = p + geom_histogram() +
  geom_vline(xintercept = score.LOO.slope$mse, colour="red") +
  xlab("Mean Squared Error (MSE)")
p

# compute p-value
p_val = sum(c(score.LOO.slope$mse,null_mse)<=score.LOO.slope$mse)/(nperm+1)
p_val
```

# Make Figure 1

```{r, warning=FALSE, message=FALSE}
fontSize_fig1 = 10

p1a = p_early_late+ ylab("aSLP")+
  theme(axis.text.x = element_text(size=fontSize_fig1),
        axis.text.y = element_text(size=fontSize_fig1),
        axis.title.x = element_text(size=fontSize_fig1),
        strip.text.x = element_text(size=fontSize_fig1),
        axis.title.y = element_text(size=fontSize_fig1))
p1b = long_plot+ ylab("aSLP")+theme(legend.text=element_text(size=fontSize_fig1))+
  theme(axis.text.x = element_text(size=fontSize_fig1),
        axis.text.y = element_text(size=fontSize_fig1),
        axis.title.x = element_text(size=fontSize_fig1),
        strip.text.x = element_text(size=fontSize_fig1),
        axis.title.y = element_text(size=fontSize_fig1))
p1c = p1+ scale_color_gradientn(colors = cols2use, name = "Slope")+theme(legend.text=element_text(size=fontSize_fig1))+
  theme(axis.text.x = element_text(size=fontSize_fig1),
        axis.text.y = element_text(size=fontSize_fig1),
        axis.title.x = element_text(size=fontSize_fig1),
        strip.text.x = element_text(size=fontSize_fig1),
        axis.title.y = element_text(size=fontSize_fig1))
p1d = p_sd_early_late+ 
  theme(axis.text.x = element_text(size=fontSize_fig1),
        axis.text.y = element_text(size=fontSize_fig1),
        axis.title.x = element_text(size=fontSize_fig1),
        strip.text.x = element_text(size=fontSize_fig1),
        axis.title.y = element_text(size=fontSize_fig1))

p_final = (p1a + p1c)/(p1b + p1d)
ggsave(filename =  here("plots","fig1.pdf"), plot=p_final, width = 10, height = 5)
```

# Descriptive stats on demographic and basic clincial variables

```{r, warning=FALSE, message=FALSE}
# add in clinical data --------------------------------------------------------
# read in lw_report
lw_report = read.csv("/Users/mlombardo/Dropbox/ACE_pheno/data/LWReport_04182020_for_ML.csv",na.strings = "NULL")
# subjects to look for in lw_report
subids2lookfor = as.character(exprPhenoData$subjectId)
# names of columns I'll look at in lw_report
vine_vars = c("vine_agemo","vine_ComTotal_DomStd","vine_DlyTotal_DomStd","vine_SocTotal_DomStd","vine_MtrTotal_DomStd","vine_AdapBehav_DomStd")
mull_vars = c("mullen_ageMo","mullen_VRT","mullen_FMT","mullen_RLT","mullen_ELT","mullen_ELC_Std")
ados_vars = c("ados_ageMo","ados_CoSoTot","ados_RRTot","ados_CoSoTotRRTot")

# number of timepoints
n_tp = 5

# tmp colnames that need the timepoint appended to the end
vine_vars_tmp = vine_vars
mull_vars_tmp = mull_vars
ados_vars_tmp = ados_vars

# pre-allocate clin_data df
cols2use = c("subjectId", "age_tx_intake",
             ados_vars[2:length(ados_vars)], 
             mull_vars[2:length(mull_vars)], 
             vine_vars[2:length(vine_vars)])
clin_data = data.frame(matrix(nrow = length(subids2lookfor), ncol = length(cols2use)))
colnames(clin_data) = cols2use
rownames(clin_data) = subids2lookfor
clin_data$subjectId = subids2lookfor

for (isub in 1:length(subids2lookfor)){

  # grab the row for the current subject
  curr_sub = subids2lookfor[isub]
  tmp_data = subset(lw_report, is.element(lw_report$subjectid,curr_sub))
  
  # find tx intage age
  tx_age_intake = tx_data[is.element(tx_data$subjectId,curr_sub),"age_tx_intake"]
  clin_data[curr_sub,"age_tx_intake"] = tx_age_intake
  
  for (i_tp in 1:n_tp){
    
    # get column time-point name
    for (ivar in 1:length(vine_vars)){
      vine_vars_tmp[ivar] = sprintf("%s_%d",vine_vars[ivar],i_tp)
    } # for (ivar in 1:length(vine_vars)){
    for (ivar in 1:length(mull_vars)){
      mull_vars_tmp[ivar] = sprintf("%s_%d",mull_vars[ivar],i_tp)
    } # for (ivar in 1:length(mull_vars)){
    for (ivar in 1:length(ados_vars)){
      ados_vars_tmp[ivar] = sprintf("%s_%d",ados_vars[ivar],i_tp)
    } # for (ivar in 1:length(ados_vars)){
    
    # Check Vineland
    data2check = tmp_data[,vine_vars_tmp]
    age_tmp = floor(data2check[1])
    if (dim(data2check)[1]!=0){
      if ((age_tmp<=tx_age_intake) & !is.na(age_tmp)){
        clin_data[curr_sub, vine_vars[2:length(vine_vars)]] =  data2check[2:length(data2check)]
      } # if (tx_age_intake<=age_tmp){
    } # if (dim(data2check)[1]!=0){

    # Check Mullen
    data2check = tmp_data[,mull_vars_tmp]
    age_tmp = floor(data2check[1])
    if (dim(data2check)[1]!=0){
      if ((age_tmp<=(tx_age_intake+1)) & !is.na(age_tmp)){
        clin_data[curr_sub, mull_vars[2:length(mull_vars)]] =  data2check[2:length(data2check)]
      } # if (tx_age_intake<=age_tmp){
    } # if (dim(data2check)[1]!=0){
    
    # Check ADOS
    data2check = tmp_data[,ados_vars_tmp]
    age_tmp = floor(data2check[1])
    if (dim(data2check)[1]!=0){
      if ((age_tmp<=tx_age_intake) & !is.na(age_tmp)){
        clin_data[curr_sub, ados_vars[2:length(ados_vars)]] =  data2check[2:length(data2check)]
      } # if (tx_age_intake<=age_tmp){
    } # if (dim(data2check)[1]!=0){
    
  } # for (i_tp in 1:n_tp){
  
} # for (isub in 1:length(subids2lookfor)){

clinvars2use = c(vine_vars[2:length(vine_vars)], mull_vars[2:length(mull_vars)], ados_vars[2:length(ados_vars)])
clin_data2add = scale(clin_data[,3:ncol(clin_data)])
data2use = cbind(data2use, clin_data2add[,clinvars2use])

# clin_vars = colnames(clin_data2add)

clin_vars = c("ados_CoSoTot","ados_RRTot","ados_CoSoTotRRTot","mullen_VRT","mullen_FMT","mullen_RLT","mullen_ELT","mullen_ELC_Std","vine_ComTotal_DomStd","vine_DlyTotal_DomStd","vine_SocTotal_DomStd","vine_MtrTotal_DomStd","vine_AdapBehav_DomStd")

df2use = merge(clin_data[,c("subjectId",clin_vars)],lw_report[,c("subjectid","gender")], by.x = "subjectId",by.y = "subjectid")
df2use = merge(df2use, tx_data_sub[,c("subjectId","age_tx_intake")], by = "subjectId")

describeBy(df2use)

table(df2use$gender)
```

# Spit out correlations between treatment slope and each clinical variable

```{r, warning=FALSE, message=FALSE}
clin_vars = c("ados_CoSoTot","ados_RRTot","ados_CoSoTotRRTot","mullen_VRT","mullen_FMT","mullen_RLT","mullen_ELT","vine_ComTotal_DomStd","vine_DlyTotal_DomStd","vine_SocTotal_DomStd","vine_MtrTotal_DomStd","vine_AdapBehav_DomStd")

clin_var_names = c("ADOS SC", "ADOS RRB", "ADOS Total", 
                   "MSEL VR", "MSEL FM", "MSEL RL", "MSEL EL",
                   "VABS Comm","VABS DL","VABS Soc","VABS Motor","VABS ABC")
cor_res = data.frame(matrix(nrow = length(clin_vars), ncol = length(c("ClinVar","r", "pval"))))
colnames(cor_res) = c("ClinVar","r", "pval")
rownames(cor_res) = clin_vars

# print out the correlation of each clinical measure separately with treatment slope
for (i in 1:length(clin_vars)){
  tmp_res = cor.test(exprPhenoData$rfx_slope, clin_data2add[,clin_vars[i]])
  cor_res[i,"ClinVar"] = clin_var_names[i]
  cor_res[i,"r"] = tmp_res$estimate
  cor_res[i,"pval"] = tmp_res$p.value
} # for (i in 1:length(clin_vars)){
cor_res$fdr_q = p.adjust(cor_res$pval, method = "fdr")
cor_res

cor_res$logp = -log10(cor_res$pval)
# fontSize=15
p = ggplot(data = cor_res, aes(x = ClinVar, y = r, fill = logp)) + 
  geom_bar(stat = "identity", colour = "black") + xlab("Clinical Measure") + ylab("Pearson's r") +
  geom_hline(yintercept = 0.31, linetype="dashed") +
  geom_hline(yintercept = -0.31, linetype="dashed") +
  ylim(-0.6,0.6) + 
  scale_fill_gradientn(colors = colorRampPalette(c("white","red"))(100), name = "-log10(p)") +
  coord_flip() +
  theme(axis.text.x = element_text(size=fontSize),
        axis.text.y = element_text(size=fontSize),
        axis.title.x = element_text(size=fontSize),
        strip.text.x = element_text(size=fontSize),
        axis.title.y = element_text(size=fontSize))
ggsave(filename = here("plots","clinmeasure_correlation_plot2.pdf"))
clin_plot = p
p
```


# Run LASSO to predict subject-specific slopes with ONLY pre-treatment clinical data 

```{r, warning=FALSE, message=FALSE}
cvfolds = 41
clin_vars = c("ados_CoSoTot","ados_RRTot","ados_CoSoTotRRTot","mullen_VRT","mullen_FMT","mullen_RLT","mullen_ELT","vine_ComTotal_DomStd","vine_DlyTotal_DomStd","vine_SocTotal_DomStd","vine_MtrTotal_DomStd","vine_AdapBehav_DomStd")

# now use the features within a LASSO
data2use = scale(clin_data[, clin_vars])

score.kCV.slope = .myLasso.CV.new.clin2(inputData = t(data2use), 
                                  inputTrait = exprPhenoData,
                                  cvfold = cvfolds,
                                  dataType = dataType,
                                  clin_vars = clin_vars)

# correlation betwwen observed and predicted slopes
cor(score.kCV.slope$data2plot)

# print out mean squared error from this model
score.kCV.slope$mse

df2plot = data.frame(predSlope = score.kCV.slope$data2plot$predVal, 
                     obsSlope = score.kCV.slope$data2plot$inputTrait, 
                     subjectId = exprPhenoData$subjectId, 
                     subgrp =exprPhenoData$subgrp)

cols2use = colorRampPalette(c("blue","red"))(100)
p2 = ggplot(data = df2plot, aes(x = predSlope, y = obsSlope)) + 
  geom_smooth(method=lm, col="black") + 
  geom_point(aes(colour=obsSlope),size=2) + scale_color_gradientn(colors=cols2use) +
  xlab("Predicted Slope") + ylab("Actual Slope")+
  theme(axis.text.x = element_text(size=fontSize),
        axis.text.y = element_text(size=fontSize),
        axis.title.x = element_text(size=fontSize),
        strip.text.x = element_text(size=fontSize),
        axis.title.y = element_text(size=fontSize))
ggsave(filename = here("plots","clin_lasso_loo_plot2a.pdf"))

p_final = single_sub_plot + p2 + plot_layout(nrow=1,ncol=2)
ggsave(filename = here("plots","clin_lasso_loo_plot2b.pdf"))
p_final

# find the relevant clinical variables
score.kCV.slope.clinALL = .myLasso.CV.genes.new.clin2(inputData = t(data2use),
                                                 inputTrait = exprPhenoData,
                                                 cvfold = cvfolds,
                                                 dataType = dataType,
                                                 clin_vars = clin_vars)

clinALL.keep.kCVslope = as.data.frame(table(unlist(score.kCV.slope.clinALL[1,])))
clinALL.keep.kCVslope = clinALL.keep.kCVslope[with(clinALL.keep.kCVslope, order(-Freq)),]
clinFilter = as.character(clinALL.keep.kCVslope[,1])
res = clinALL.keep.kCVslope
colnames(res)[1] = "ClinVar"
write.csv(res, file = here("results",sprintf("res_lasso_%dfold_bestclinvarsALL_Microarray2.csv",cvfolds)))
res

# permutation test on MSE
null_mse = .myRandFn.CV.new.clin2(inputData = t(data2use),
                            inputTrait = exprPhenoData,
                            nperm = nperm,
                            cvfold = cvfolds,
                            dataType = dataType,
                            seed2use=123,
                            verbose=FALSE,
                            clin_vars = clin_vars)

# make histogram plot of MSE null distribution
df2plot = data.frame(mse = null_mse)
p = ggplot(data = df2plot, aes(x = mse))  
p = p + geom_histogram() + 
  geom_vline(xintercept = score.kCV.slope$mse, colour="red") +
  xlab("Mean Squared Error (MSE)")
p  

# compute p-value
p_val = sum(c(score.kCV.slope$mse,null_mse)<=score.kCV.slope$mse)/(nperm+1)
p_val
```

# Make Figure 2

```{r, warning=FALSE, message=FALSE}
fontSize_fig2 = 10

p2a = p2+ scale_color_gradientn(colors = cols2use, name = "Slope") + theme(legend.text=element_text(size=fontSize_fig2))+
  theme(axis.text.x = element_text(size=fontSize_fig2),
        axis.text.y = element_text(size=fontSize_fig2),
        axis.title.x = element_text(size=fontSize_fig2),
        strip.text.x = element_text(size=fontSize_fig2),
        axis.title.y = element_text(size=fontSize_fig2))
p2b = clin_plot+ theme(legend.text=element_text(size=fontSize_fig2))+
  theme(axis.text.x = element_text(size=fontSize_fig2),
        axis.text.y = element_text(size=fontSize_fig2),
        axis.title.x = element_text(size=fontSize_fig2),
        strip.text.x = element_text(size=fontSize_fig2),
        axis.title.y = element_text(size=fontSize_fig2))

p_final = (p2a + p2b)
ggsave(filename =  here("plots","fig2.pdf"), plot=p_final, width = 10, height = 5)
```